|

### 2022-01-17 text/ManuscriptDraft.Rmd
Manuscript update rev1

---
new_session: FALSE
title: |
  **Monitoring the phenology of individual flowers using automatic tracking and filtering**
author:
- Hjalte M. R. Mann
- Alexandros Iosifidis
- Toke T. Høye
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
  html_document:
    df_print: paged
fontfamily: mathpazo
fontsize: 12pt
geometry: margin = 1in
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage[left]{lineno}
- \linenumbers
- \usepackage{dcolumn}
- \usepackage{caption}
- \usepackage{float}
- \usepackage{afterpage}
- \usepackage{siunitx}
- \usepackage{amsmath}
keywords: Tracking, flowering phenology, Arctic, DBSCAN
bibliography: ./library.bib
csl: ./journal-of-ecology.csl
abstract: ABSTRACT | Often simple variables will be used to describe the flowering phenology of a population of plants, e.g. onset or peak of flowering. Here we show that image-based monitoring of field plots at very high temporal resolution can return information on flowering phenology at the level of indiviuals. Further, we present a framework for automatically tracking, filtering, and evaluating flowers in time-lapse image series. We compare the results of automatic tracking with manual tracking.
---

@@

For the individual plant, timing of flowering is of utmost importance. Precocious flowering means that the plant has failed to exploit the whole temporal window for accumulating resources before allocating energy to flowering. Oppositely, flowering too late limits the time for reproduction before the end of the growing season [@ELZINGA2007]. Further, flowering may need to be synchronous with pollinator activity for successful reproduction. Flowering phenology may plastically change as a response to abiotic cues in the environment, such as timing of spring, temperature, and photoperiod, but variation in flowering phenology is partly heritable and shaped by selective forces from the abiotic and biotic environment and thus sensitive to climate change.

Monitoring of flower phenology at high temporal resolution is laboursome and time-consuming, particularly in logistically challenging environments such as the Arctic. Consequently, simple variables are often used as proxies for the flowering phenology of a population, such as the date for onset of flowering, often derived from weekly observations of sample plots. Such proxies may fail to reveal dynamics in flowering phenology for example caused by changes in climate.

Automatic image-based monitoring of flowering phenology can return phenology data for specific species at very high temporal resolution (Mann et al., in prep), but phenological responses at the individual level may be indiscernible regardless of the temporal resolution of the data at population level. For example, a shortening of individual flower longevity may not be directly obvious at the population level. In fact, many research questions can only be explored on the basis of individual phenology data. For example, investigating the association between reproductive success and timing of flowering or flower longevity requires phenology data at the level of individuals. Similarly so for investigating whether flower visitation rates and/or reproductive success depends on the timing of flowering for the individual flower.

@@

We present and evaluate an automatic flower tracking and filtering algorithm. Many tracking algorithms, such as the Simple online and realtime tracking (SORT) [@sort], use Kalman filtering to predict the future location of an object by estimating its velocity [@kalmann]. However, Kalman filtering assumes linear motion of the object and does not handle abrupt motion well. As the flowers are constrained in movement by their stalk, their movement does not follow the assumption of linear motion. We therefore base the track assignment on the minimum distance between centroids of bounding boxes. However, we introduce three user-set parameters that can help optimise tracking performance.

@@

Tracking individual flowers enables the possibility of assigning reproductive success to the individual, for example by observation of seed set. Thereby it could be explored whether reproductive success is affected by timing and length of flowering. By simultaneously tracking flower visits, these could be assigned to the individual flower and visitation rates per flower could be calculated per flower and related to reproductive success. Further, any information of taxonomic grouping could refine this analyses. We argue that the possibility of tracking phenology at the level of individuals will facilitate investigation of a novel set of research questions and is an important contribution to the shift towards efficient and automatic monitroing of ecological processes.

@@

`r word_comment("Text below is maybe more suited for intro/disc? Attempting to explain the reasoning of the method.", "The")` tracking algorithm tracks objects based on distances between centroids of bounding boxes. The algorithm has a set of user adjusted parameters that can optimize tracking accuracy. The parameters are particularly relevant for optimal tracking of objects that are constrained to a specific area such as flowers. It is important to note, however, that the tracking algorithm could be used to track any objects. The algorithm can be applied both offline (on a set of detections/annotations that have already been produced) or online (real-time tracking frame per frame). The speed of the tracking algorithm depends on the computational power available as well as the number of objects that are being tracked. The method is fast, however. Tracking of a series containing 85 objects ran at ~0.02 seconds per frame on a standard laptop.

>>>>

---
new_session: FALSE
title: |
  **Towards individual-based pollination ecology: Automatic tracking of life histories of individual flowers**
author:
- Hjalte M. R. Mann
- Alexandros Iosifidis
- Toke T. Høye
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
  html_document:
    df_print: paged
fontfamily: mathpazo
fontsize: 12pt
geometry: margin = 1in
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage[left]{lineno}
- \linenumbers
- \usepackage{dcolumn}
- \usepackage{caption}
- \usepackage{float}
- \usepackage{afterpage}
- \usepackage{siunitx}
- \usepackage{amsmath}
keywords: Tracking, flowering phenology, Arctic, DBSCAN
bibliography: ./library.bib
csl: ./journal-of-ecology.csl
abstract: ABSTRACT |

A fundamental problem in pollination ecology is that plant reproduction consequences of pollinator visits to a flower throughout its lifespan is very difficult to track. Often simple statistics are used to describe the flowering phenology of a population of plants, e.g. onset or peak of flowering. Similarly, flower visitation rates are based on crude observation of flower populations. Here we show that image-based monitoring of field plots at very high temporal resolution can return information on flowering phenology at the level of indiviuals. Further, we present a framework for automatically tracking, filtering, and evaluating flowers in time-lapse image series. We compare the results of automatic tracking with manual tracking. Individual flower tracking could facilitate finegrained studies on plant reproductive success.
---

@@

For the individual plant, timing of flowering is of utmost importance. Precocious flowering means that the plant has failed to exploit the whole temporal window for accumulating resources before allocating energy to flowering. Oppositely, flowering too late limits the time for reproduction before the end of the growing season [@ELZINGA2007]. Both precocious and late flowering increase the risk of frost damage to the flowers. Further, flowering may need to be synchronous with pollinator activity for successful reproduction. Flowering phenology may plastically change in response to abiotic cues in the environment, such as timing of spring, temperature, and photoperiod, but flowering phenology is partly heritable and shaped by selective forces from the abiotic and biotic environment and in any case sensitive to climate change.

Monitoring flower phenology at high temporal resolution is time-consuming, particularly in logistically challenging environments such as the Arctic. Consequently, flowering phenology of a population is often characterized simply by the date for onset of flowering, typically derived from weekly observations of sample plots [@Prevey2021]. Such inaccurate estimates may fail to reveal dynamics in flowering phenology for example caused by changes in climate.

Automatic image-based monitoring of flowering phenology can return phenology data for specific species at very high temporal resolution (Mann et al., in review), but phenological responses at the individual level may be indiscernible regardless of the temporal resolution of the data at population level. For example, a shortening of individual flower longevity may not be directly obvious at the population level. In fact, many research questions can only be explored on the basis of individual phenology data. For example, investigating the association between reproductive success and timing of flowering or flower longevity requires phenology data at the level of individuals. Similarly so for investigating whether flower visitation rates and/or reproductive success depends on the timing of flowering for the individual flower.

@@

We present and evaluate an automatic flower tracking and filtering algorithm. Many tracking algorithms, such as the Simple online and realtime tracking (SORT) [@sort], use Kalman filtering to predict the future location of an object by estimating its velocity [@kalmann]. However, Kalman filtering assumes linear motion of the object and does not handle abrupt motion well. As the flowers are constrained in movement by their stalk, their movement does not follow the assumption of linear motion. We therefore base the track assignment on the minimum distance between centroids of bounding boxes and introduce three user defined parameters that can help optimise tracking performance.

@@

Our method for tracking individual flowers facilitates the possibility of assigning indicators of reproductive success and factors driving variation in seed set to the individual. For instance, it could be explored whether seed set is affected by timing and length of flowering and/or abiotic conditions such as frost events. Hereby, the longstanding question of what drives variation in reproductive success of individual flowers could be explored. By simultaneously tracking flower visits, these could be assigned to the individual flower and visitation rates per flower could be calculated per flower and related to reproductive success. Further, any information of taxonomic grouping could facilitate comparative studies on pollination effectiveness between taxonomic groups of pollinators. We argue that the possibility of tracking phenology at the level of individuals will facilitate investigation of a novel set of research questions and is an important contribution to the shift towards efficient and automatic monitoring of ecological processes.

@@

The tracking algorithm tracks objects based on distances between centroids of bounding boxes. The algorithm has a set of user adjusted parameters that can optimize tracking accuracy. The parameters are particularly relevant for optimal tracking of objects that are constrained to a specific area such as flowers. It is important to note, however, that the tracking algorithm could be used to track any objects. The algorithm can be applied both offline (on a set of detections/annotations that have already been produced) or online (real-time tracking frame per frame). The speed of the tracking algorithm depends on the computational power available as well as the number of objects that are being tracked. The method is fast, however. Tracking of a series containing 85 objects ran at ~0.02 seconds per frame on a standard laptop.
